Higher-Order unification for free!

### 2024-07-02 paper/main.tex
eta-beta -> beta-eta

  The object logic we are interested in is
  Coq's type theory~\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure %for it 
  using the meta language
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\eta\beta$-equivalence and features a
  higher-order unification procedure \Ue 
  %restricted to the
  for the
  pattern fragment~\cite{miller92jsc}.
  Elpi offers an encoding of Coq terms that is suitable
  for meta programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but that restricts \Ue to first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose to translate \Fo{} unification problems
  into an alternative encoding called \Ho, from which we derive \Uo,
  the higher-order unification procedure of \Fo{}. \Uo
  honours $\eta\beta$-equivalence for terms
  within the pattern fragment, and allows the use for 
  heuristics when unification is performed with terms outside the pattern fragment.
  Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.

@@

The object logic we are interested in is Coq's type theory. We aim to develop a
higher-order unification-based proof search procedure using the meta language
Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
$\eta\beta$-equivalence and features a higher-order unification procedure \Ue
for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
for meta programming but that restricts \Ue to first-order unification problems
only. We refer to this basic encoding as \Fo.

In this paper we propose a more well-behaved encoding called \Ho, and show how
to translate unification problems from \Fo{} to corresponding ones in \Ho.
Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
that honours $\eta\beta$-equivalence (for Coq functions), addresses problems
within the pattern fragment, and allows for the use of heuristics to deal with
problems outside the pattern fragment. Moreover, as \Uo{} delegates most of the
work to \Ue, it can be used to efficiently simulate a logic program in \Fo{} by
taking advantage of unification-related optimizations of the meta language, such
as clause indexing.
\end{comment}

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta-language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho, the language of the meta language, and we show that
the higher-order unification procedure of the meta language \Ue{} can be
used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

@@

% In this section we describe a basic compilation scheme that we refine
% later, in the following sections. This scheme is sufficient to implement
% a \hstep that respects $\beta$-conversion for terms in \llambda.
% The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
% the support for terms outside \llambda in \cref{sec:beta}.

>>>>

  The object logic we are interested in is
  Coq's type theory~\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure %for it 
  using the meta language
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\beta\eta$-equivalence and features a
  higher-order unification procedure \Ue 
  %restricted to the
  for the
  pattern fragment~\cite{miller92jsc}.
  Elpi offers an encoding of Coq terms that is suitable
  for meta programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but that restricts \Ue to first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose to translate \Fo{} unification problems
  into an alternative encoding called \Ho, from which we derive \Uo,
  the higher-order unification procedure of \Fo{}. \Uo
  honours $\beta\eta$-equivalence for terms
  within the pattern fragment, and allows the use for 
  heuristics when unification is performed with terms outside the pattern fragment.
  Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.

@@

The object logic we are interested in is Coq's type theory. We aim to develop a
higher-order unification-based proof search procedure using the meta language
Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
$\beta\eta$-equivalence and features a higher-order unification procedure \Ue
for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
for meta programming but that restricts \Ue to first-order unification problems
only. We refer to this basic encoding as \Fo.

In this paper we propose a more well-behaved encoding called \Ho, and show how
to translate unification problems from \Fo{} to corresponding ones in \Ho.
Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
that honours $\beta\eta$-equivalence (for Coq functions), addresses problems
within the pattern fragment, and allows for the use of heuristics to deal with
problems outside the pattern fragment. Moreover, as \Uo{} delegates most of the
work to \Ue, it can be used to efficiently simulate a logic program in \Fo{} by
taking advantage of unification-related optimizations of the meta language, such
as clause indexing.
\end{comment}

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta-language Elpi encompasses
$\beta\eta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons automatic proof search procedures typically employ a unification
procedure that only captures a $\beta\eta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho, the language of the meta language, and we show that
the higher-order unification procedure of the meta language \Ue{} can be
used to simulate a higher-order unification procedure \Uo for
the object language that features $\beta\eta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

@@

% In this section we describe a basic compilation scheme that we refine
% later, in the following sections. This scheme is sufficient to implement
% a \hstep that respects $\beta$-conversion for terms in \llambda.
% The extension to $\beta\eta$-conversion is described in \cref{sec:eta} and
% the support for terms outside \llambda in \cref{sec:beta}.
